---
title: "Species Distribution Modeling of *Phyllomedusa distincta*"
output: gfm
editor: visual
knitr:
  opts_chunk:
    warning: false
    message: false
bibliography: sdm.bib
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(terra)
library(spThin)
library(janitor)
library(here)
library(rasterVis)
library(tidyverse)
library(patchwork)
library(usdm)
library(tidyterra)
library(rnaturalearth)
library(ENMeval)
```

# Current

## Variable selection

Bioclims:

I'm using CHELSA V2.1 CMIP6 Bioclims at 30 arc-seconds resolution, aggregated to 5 arc-minutes @karger2017. **Downloaded on 2024-04-27.**

For projecting to past climates, I'm using [CHELSA TraCE21k v1.0 Bioclims](https://cp.copernicus.org/preprints/cp-2021-30/cp-2021-30.pdf) at 1000-year intervals back to the LGM. 30 arc-seconds resolution, aggregated to 5 arc-minutes @karger2023. **Download on 2024-04-27.**

## Variable processing

First, I'm reading in the localities and plotting them to make sure there isn't anything wild. Everything looks reasonable!

```{r}
#| label: locs
#| message: false

  
locs_sf <- read_csv(here("distincta_localities.csv")) %>%
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326,
           remove = FALSE)

# atlantic forest shapefile for plotting
af <- read_sf(here("climate", "atlantic_forest.geojson")) %>% 
  terra::vect() %>% 
  terra::aggregate() %>% 
  st_as_sf()

# plot
ggplot() +
  geom_sf(data = st_geometry(af), fill = "gray") +
  geom_sf(data = locs_sf, shape = 21, fill = "darkgreen") +
  coord_sf(xlim = c(-55, -45), ylim = c(-31, -22)) +
  theme_minimal()
```

Creating a convex hull with an 2 degree buffer.

```{r}
#| label: mcp-all
#| warning: false

mcp_all <- st_convex_hull(st_union(locs_sf)) %>%
  st_buffer(dist = units::set_units(2, degree)) 

ggplot() +
  geom_sf(data = af, fill = "gray") +
  geom_sf(data = mcp_all, fill = "transparent") +
  geom_sf(data = locs_sf, shape = 21, fill = "darkgreen") +
  coord_sf(xlim = c(-55, -45), ylim = c(-31, -22)) +
  theme_minimal()
```

Read in bioclims.

```{r}
#| label: crop-env
#| warning: false

land <- rnaturalearth::ne_countries(scale = "large", continent = "South America") %>%
  vect() %>% 
  terra::project("epsg:4326")

bioclims <- terra::rast(list.files(here("climate"), pattern = "_0kyBP.tif", full.names = TRUE)) %>% 
  terra::project("epsg:4326") %>% 
  # crop the bioclims to the coastline
  crop(land) %>% 
  mask(land)

# get the only the bioclim labels
names(bioclims) <- str_extract(names(bioclims), "bio[\\d+]*")

ggplot() +
  geom_spatraster(data = bioclims$bio1) +
  scale_fill_whitebox_c("muted") +
  geom_sf(data = af, fill = "transparent") +
  geom_sf(data = mcp_all, fill = "transparent") +
  geom_sf(data = locs_sf, shape = 21, fill = "darkgreen") +
  coord_sf(xlim = c(-55, -45), ylim = c(-31, -22)) +
  labs(fill = "Average\nAnnual Temp. (C°)") +
  theme_minimal()

```

## General SDM steps {.tabset}

I'm going to perform the following steps:

-   spatially thin localities\
-   extract background environmental data (10,000 points max) for predictor correlation and modeling
-   remove predictors w/ high variance inflation factor (10)
-   Use Maxent for modeling
    -   L, LQ, H, LQH feature classes
    -   regularization multipliers from 0.5 to 5.0 in 0.5 increments to test a wide range of regularization
    -   spatial cross validation (block) for model selection
    -   select models first by AICc (model fit), followed by ommission error rate (prediction)

#### Spatial thin

Spatial thin. I'm using a 20 km buffer to reduce the impact of spatial autocorrelation.

```{r}
#| label: spthin

set.seed(39434)

#run spthin algorithm. This returns 100 possible combinations of removed localities
output <-
  spThin::thin(
    locs_sf,
    'latitude',
    'longitude',
    'species',
    thin.par = 20,
    reps = 100,
    locs.thinned.list.return = TRUE,
    write.files = FALSE,
    write.log.file = FALSE,
    verbose = FALSE
  )

# I want to maximize the # of localities returned  
maxThin <- which(sapply(output, nrow) == max(sapply(output, nrow)))

# if there are multiple iterations with the max # of localities, pick one
maxThin <-
  output[[ifelse(length(maxThin) > 1, maxThin[1], maxThin)]]

# subset locs to match only thinned locs
locs_sf <- locs_sf[as.numeric(rownames(maxThin)), ]

# get the unused locs as a testing data set
# this probably isn't useful since they will overlap heavily with the training data, but seems like another piece of info to look at
test <- locs_sf[-as.numeric(rownames(maxThin)), ]

ggplot() + 
  geom_sf(data = af, fill = "gray") + 
  geom_sf(data = locs_sf, shape = 21, fill = "darkgreen") +
  coord_sf(xlim = c(-55, -45), ylim = c(-31, -22)) +
  theme_minimal()
```

Write thinned locs to file.

```{r}
#| label: write-thinned
#| eval: false

st_write(locs_sf, here("distincta_thinned.geojson"))

```

#### Predictor VIF

Sample 10000 background points. Only returned 7247 points.

```{r}
#| label: bg
set.seed(1938)

# for variable correlations
bg_envt <- terra::spatSample(bioclims, 10000, 
                               warn=TRUE, 
                               na.rm = TRUE, 
                               as.df = TRUE,
                               xy = TRUE)

# for use in ENMevaluate
bg_coords <- bg_envt[,c("x", "y")]
```

Next, I'll extract the values for the background points and perform variance inflation factor stepwise selection with a VIF threshold of 10.

```{r}
#| label: vif
# extract values
bg_corr <- bg_envt %>% select(-x, -y)

usdm::vifstep(bg_corr, th=10)
```

The final variable list: BIO2, BIO4, BIO5, BIO8, BIO9, BIO13, BIO19

```{r}
# label: pred

predictors <- bioclims[[c("bio2", "bio4", "bio5", "bio8", "bio9", "bio13", "bio19")]]
```

#### Maxent model

I'm using a spatial cross-validation (block) evaluation approach to reduce the impact of spatial autocorrelation on inference. Following @radosavljevic2014.

```{r}
#| label: folds
set.seed(70787)
coords <- st_coordinates(locs_sf)
colnames(coords) <- c("x", "y")

folds <- ENMeval::get.block(occs = coords, bg = bg_coords, orientation = "lon_lat")
```

Plot the folds assignments of localities.

```{r}
#| label: plot-folds

locs_folds <- locs_sf %>% 
  mutate(fold = as.factor(folds$occs.grp))

ggplot() +
  geom_sf(data = af, fill = "gray") +
  geom_sf(data = locs_folds, aes(fill = fold), shape = 21) +
  scale_fill_viridis_d() +
  coord_sf(xlim = c(-55, -45), ylim = c(-31, -22)) +
  theme_minimal()
  
```

Run the model. Predictions are clamped to prevent extrapolation.

```{r}
#| label: run-model
#| eval: false

set.seed(34223)

# the vector of regularization multipliers to test
rms <- seq(0.5, 5, 0.5)

# convert the terra object to a raster stack for use in EMNeval
predictors_stack <- raster::stack(predictors)

# iterate model building over all chosen parameter settings
sdm <-
  ENMeval::ENMevaluate(
    occs = coords,
    envs = predictors_stack,
    bg = bg_coords,
    tune.args = list(fc = c("L", "LQ", "H", "LQH"), rm = rms),
    partitions = "user",
    algorithm = "maxent.jar",
    # clamping to prevent model extrapolation
    doClamp = TRUE,
    user.grp = folds,
    taxon.name = "distincta",
    n.bg = 10000,
    parallel = TRUE,
    numCores = 6
  )
```

```{r}
#| label: write-model
#| eval: false

# write the model to file
write_rds(sdm, here("output", "sdm.rds"))
```

```{r}
#| label: read-model
#| echo: false

sdm <- read_rds(here("output", "sdm.rds"))
```

##### Model evaluation

Let's take a look at the model results.

```{r}
#| label: mod-results
eval_table <- sdm@results
eval_mods <- sdm@models

names(eval_mods) <-
  str_replace_all(names(eval_mods), "\\.", "\\_")
```

Select the final model. I'm going to take a look at tables of delta AICc, omission rate, and AUC to see how close the models are.

Four models are within 2 AICc of each other. I selected the model with LQ feature class and a regularization multiplier of 2 because this is the simplest model out of the three (6 coefficients), contains the lowest omission rate (both 10th percentile and MTP), and has the highest AUC performance on the validation set.

```{r}
#| label: eval-table

eval_table %>% 
  select(delta.AICc, AICc, or.10p.avg, or.mtp.avg, auc.diff.avg, auc.val.avg, rm, fc, ncoef) %>%
  arrange(delta.AICc) %>% 
  head(10) %>% 
  knitr::kable()
```

Select model.

```{r}
#| label: select-model
mod <- eval_mods$fc_LQ_rm_2
opt_seq_tune <- "fc.LQ_rm.2"
```

Plot the variable contributions and response curves. The most important variables are bio2, bio13, and bio9. All precipitation!

```{r}
#| label: var-importance
plot(mod)
```

```{r}
#| label: write-var-imp
#| echo: false
#| eval: false
png(here("output", "plots", "variable_contributions.png"), width = 15, height = 12, units = "cm", res = 600)
plot(mod)
dev.off()
```

Suitability increases with precipitation amount of the wettest month, decreases with precipitation seasonality, and increases with mean monthly precipitation of the warmest quarter.

```{r}
#| label: resp-curve
dismo::response(mod)
```

```{r}
#| label: write-resp-curve
#| echo: false
#| eval: false
png(here("output", "plots", "response_curves.png"), width = 20, height = 12, units = "cm", res = 600)
dismo::response(mod)
dev.off()


```

##### Project

I'm projecting the model to the study area extent.

```{r}
#| label: proj-cur
pred <- ENMeval::eval.predictions(sdm)[[opt_seq_tune]] %>% 
  rast()


ggplot() +
  geom_spatraster(data = pred) +
  scale_fill_whitebox_c("muted") +
  geom_sf(data = af, fill = "transparent") +
  geom_sf(data = locs_sf, fill = "darkgreen", shape = 21) +
  coord_sf(xlim = c(-55, -45), ylim = c(-31, -22)) +
  theme_minimal()
```

```{r}
#| label: proj-cur-thresh

pred_thresh <- pred[[opt_seq_tune]]
# observed suitabilities
obs_suit <- terra::extract(pred_thresh, locs_sf)
# minimum training presence
min_suit <- min(obs_suit)

pred_thresh[pred_thresh >= min_suit] <- 1
pred_thresh[pred_thresh < min_suit] <- 0

ggplot() +
  geom_spatraster(data = pred_thresh) +
  scale_fill_whitebox_c("muted") +
  geom_sf(data = af, fill = "transparent") +
  geom_sf(data = locs_sf, fill = "darkgreen", shape = 21) +
  coord_sf(xlim = c(-55, -45), ylim = c(-31, -22)) +
  theme_minimal()
```

## References

::: refs
:::
